Pretty darn good control: when are approximate solutions better than approximate models

### 2023-07-05 manuscript/manuscript.tex
added tex

% !TeX program = pdfLaTeX
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}

% common operators
\newcommand*{\der}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}

% structural markers
\def\div{\;|\;}
\def\bigdiv{\;\big|\;}
\def\Bigdiv{\;\Big|\;}
\def\Biggdiv{\;\Bigg|\;}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

%% load any required packages here

@@

\begin{document}

\title{Title here \thanks{Grants or other notes about the article that
should go on the front page should be placed here. General
acknowledgments should be placed at the end of the article.} }
 \subtitle{Do you have a subtitle? If so, write it here}

    \titlerunning{Short form of title (if too long for head)}

\author{  Äüthör 1 \and  Âuthóř 2 \and  }

    \authorrunning{ Short form of author list if too long for running
head }

\institute{
        Äüthör 1 \at
     Department of YYY, University of XXX \\
     \email{\href{mailto:abc@def}{\nolinkurl{abc@def}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Âuthóř 2 \at
     Department of ZZZ, University of WWW \\
     \email{\href{mailto:djf@wef}{\nolinkurl{djf@wef}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
The text of your abstract. 150 -- 250 words.
\\
\keywords{
        key \and
        dictionary \and
        word \and
    }

    \subclass{
                    MSC code 1 \and
                    MSC code 2 \and
            }

\end{abstract}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

@@

Your text comes here. Separate text sections with \cite{Mislevy06Cog}.

\hypertarget{sec:1}{%
\section{Background}\label{sec:background}}

% Text with citations by \cite{Galyardt14mmm}.
In this section we first review two well-established techniques commonly used in sustainable fishery management. 
These are the maximum sustainable yield (MSY) and the constant escapement (CE) approaches. 
After this, deep reinforcement learning is briefly reviewed

\hypertarget{sec:2}{%
\subsection{Fishery management}\label{sec:fishery}}

\hypertarget{sec:3}{%
\subsection{Deep reinforcement learning}\label{sec:rl}}

\hypertarget{sec:4}{%
\section{Dynamical models used}\label{sec:model}}

In this section we present three models of increasing complexity which plausibly describe the population dynamics of a marine ecosystem.
These models will form the test beds for the comparison between classical fishery management strategies and DRL.

\hypertarget{sec:4.1}{%
\subsection{A one-dimensional tipping point model.}\label{p:may}}

Consider a population $V$ whose dynamics is given by
\begin{align}
  \label{eq:may77}
  \der{V}{t} = rV\left(
    1-V/K
    \right) 
    -
    \frac{\beta H V^2}{V_0^2 + V^2}. 
\end{align}
This model has been used in~\cite{may77} to describe a grazing ecosystem, where a species $V$ of vegetation is harvested by a constant herbivore population $H$. 
% Here the only dynamic degree of freedom is $V$, while $r,\ K,\ \beta,\ V_0,$ and $H > 0$ are fixed positive constants.

In~\eqref{eq:may77}, a population $V$ grows logistically with rate $r$ up to carrying capacity $K$.
This is expressed by the first term in the equation,
\begin{align*}
  L(V \div r,\ K) := rV\left(1 - V / K\right).
\end{align*}
Moreover, $V$ is predated on by a (constant) population $H$, as can be seen from the negative term
\begin{align*}
  F(V,\ H \div \beta,\ V_0) := \frac{\beta H V^2}{V_0^2 + V^2}
\end{align*}
which saturates to $\beta H$ as $V\to\infty$, and whose half maximum is $V_0$, i.e.\ $F(V=V_0,\ H; \beta,\ V_0)=\beta H/2$.

Ref.~\cite{may77} studies the fixed points of~\eqref{eq:may77} in order to show that in certain parameter regimes, its dynamics can undergo a \emph{catastrophe}.
A catastrophe is a sudden change in the state of the system from one stable state to another---often, the final state is ecologically detrimental, possibly associated with extinction or near-extinction events.

Fig.~\ref{fig:tbd} shows the stable $V$ populations for differing values of $H$. 
Here one sees that as $H\to T_2$, the top stable state is annihilated with the unstable fixed point.
This way, if, e.g.\ $H$ were to slowly drift until $H=T_2$, the system would collapse to the low stable state, leading to a near extinction of $V$.

% For fixed values of $r,\ K,\ V_0$ and $\beta$, varying $H$ may produce a \emph{catastrophe} (see Fig.~\ref{fig:tbd})---a sudden disappearance of a stable fixed point of the system which .

\hypertarget{sec:5}{%
\section{Results}\label{sec:results}}

\hypertarget{sec:6}{%
\section{Discussion}\label{sec:discussion}}

\hypertarget{paragraph-headings}{%
\paragraph{Paragraph headings}\label{paragraph-headings}}

Use paragraph headings as needed. 
Use paragraph headings as needed. 
Use paragraph headings as needed. 
Use paragraph headings as needed. 
Use paragraph headings as needed. 
Use paragraph headings as needed. 
Use paragraph headings as needed.

\begin{align}
a^2+b^2=c^2
\end{align}

>>>>

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{Pretty darn good control: when are approximate solutions better
than approximate models}

\author{
    Felipe Montealegre-Mora
   \\
     \\
   \\
  \texttt{} \\
   \And
    Marcus Lapeyrolerie
   \\
     \\
   \\
  \texttt{} \\
   \And
    Melissa Chapman
   \\
     \\
   \\
  \texttt{} \\
   \And
    Abigail G. Keller
   \\
     \\
   \\
  \texttt{} \\
   \And
    Carl Boettiger
   \\
     \\
   \\
  \texttt{\href{mailto:cboettig@berkeley.edu}{\nolinkurl{cboettig@berkeley.edu}}} \\
  }

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{lineno}
\usepackage{setspace}
\usepackage{bookmark}
\usepackage{float}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tikz-cd}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\usepackage{url}
\linenumbers
\doublespacing
\begin{document}
\maketitle

\begin{abstract}
Existing methods for optimal control struggle to deal with the
complexity commonly encountered in real-world systems, including
dimensionality, process error, model bias and data heterogeneity.
Instead of tackling these system complexities directly, researchers have
typically sought to simplify models to fit optimal control methods. But
when is the optimal solution to an approximate, stylized model better
than an approximate solution to a more accurate model? While this
question has largely gone unanswered owing to the difficulty of finding
even approximate solutions for complex models, recent algorithmic and
computational advances in deep reinforcement learning (DRL) might
finally allow us to address these questions. DRL methods have to date
been applied primarily in the context of games or robotic mechanics,
which operate under precisely known rules. Here, we demonstrate the
ability for DRL algorithms using deep neural networks to successfully
approximate solutions (the ``policy function'' or control rule) in a
non-linear three-variable model for a fishery without knowing or ever
attempting to infer a model for the process itself. We find that the
reinforcement learning agent discovers a policy that outperforms both
constant escapement and constant mortality policies---the standard
family of policies considered in fishery management. This DRL policy has
the shape of a constant escapement policy whose escapement values depend
on the stock sizes of other species in the model.
\end{abstract}

\keywords{
    Optimal Control
   \and
    Reinforcement Learning
   \and
    Uncertainty
   \and
    Decision Theory
  }

@@

Much effort has been spent grappling with the complexity of our natural
world in contrast to the relative simplicity of the models we use to
understand it. Heroic amounts of data and computation are being brought
to bear on developing better, more realistic models of our environments
and ecosystems, in hopes of improving our capacity to address the many
planetary crises. But despite these efforts and advances, we remain
faced with the difficult task of figuring out how best to respond to
these crises. While simplified process models for the population
dynamics have historically allowed for exploration of large decision
spaces, the new wave of rich models are applied to highly oversimplified
descriptions of potential actions they seek to inform. For instance,
Global Circulation Models (GCMs) such as HadCM3 (Pope et al. 2000; C.
Gordon et al. 2000; Collins, Tett, and Cooper 2001) model earth's
climate using 1.5M variables, while the comparably vast potential action
space is modeled much more minimalistically, with 5 SSP socioeconomic
storylines and 7 SSP-RCP marker scenarios summarizing the action space
at the IPCC (Riahi et al. 2017).

Even as our research community develops simulations of the natural world
that fit only in supercomputers, we analyze a space of policies that
would fit on index cards. Similar combinations of rich process models
and highly simplified decision models (often not even given the status
of `model') are common. Modeling the potential action space as one of a
handful of discrete scenarios is sometimes a well justified
acknowledgement of the constraints faced by real-world decision-makers
-- particularly in the context of multilateral decisions -- and may seem
to reflect a division of responsibilities between `scientists' modeling
the `natural processes' and policy-makers who make the decisions. But,
more often, this simplification of decision choices is simply
mathematically or conceptually convenient. This simplification reflects
trade-offs between tractablity and complexity at the basis of any
mathematical modeling -- if we make both the state space and action
space too realistic, the problem of finding the best sequence of actions
quickly becomes intractable. However, emerging data-driven methods from
machine learning offer a new choice -- algorithms that can find good
strategies in previously intractable problems, but at the cost of
opacity.

In this paper, we focus on a well-developed application of model-based
management of the natural world that has long illustrated the trade-offs
between model complexity and policy complexity: the management of marine
fisheries. Fisheries management is both an important issue to society as
well as a rich and frequent test-bed of ecological management more
generally. Fisheries are an essential natural resource that provide the
primary source of protein for one in every four humans, and have faced
widely documented declines due to over-fishing Costello et al. (2016).
Fisheries management centers around the process of sampling populations
to determine fishing quotas based on population estimates. This decision
is often guided by a model of the dynamics of the system. Our paper
focuses on the decision side of this problem rather than the measurement
step.

Fisheries management has roots in both the fields of \emph{ecosystem
management} and \emph{natural resource economics}. Both fields might
trace their origins to the notion of maximum sustainable yield (MSY),
introduced independently by a fisheries ecologist (Schaefer 1954) and
the economist (H. S. Gordon and Press 1954) in the same year. From this
shared origin, each field would depart from the simplifying assumptions
of the Gordon-Schaefer model in divergent ways, leading to different
techniques for deriving policies from models.The heart of the management
problem is easily understood: a manager seeks to set quotas on fishing
that will ensure the long-term profitability and sustainability of the
industry. Mathematical approaches developed over the past century may be
roughly divided between these two fields: (A) ecologists, focused on
ever more realistic models of the biological processes of growth and
recruitment of fish while considering a relatively stylized suite of
potential management strategies, and (B) economists, focused on far more
stylized models of the ecology while exploring a far less constrained
set of possible policies. The economist's approach can be characterized
by the mathematics of a Markov decision process (MDP Colin W. Clark
1973; Colin W. Clark 1990; Marescot et al. 2013), in which the
decision-maker must observe the stock each year and recommend a possible
action. In this approach, the policy space that must be searched is
exponentially large -- for a management horizon of T decisions and a
space of N actions, the number of possible policies is \(N^T\). In
contrast, fisheries ecologists and ecosystem management typically search
a space of policies that does not scale with the time horizon. Under
methods such as ``Management Strategy Evaluation'' (MSE, (Punt et al.
2016)) a manager identifies a candidate set of ``strategies'' a priori,
and then compares the performance of each strategy over a suite of
simulations to determine which strategy gives the best outcome
(i.e.~best expected utility). This approach is far more amenable to
complex simulations of fisheries dynamics and more closely corresponds
to how most marine fishing quotas are managed today in the United States
(see stock assessments documented in RAM Legacy Stock Assessment
Database 2020).

\includegraphics[width=1\linewidth]{figures/textbox}

Recent advances in machine learning may allow us to once again bridge
these approaches, while also bringing new challenges of their own. Novel
data-driven methods have allowed these models to evolve into ever more
complex and realistic simulations used in fisheries management, where
models with over 100 parameters are not uncommon (RAM Legacy Stock
Assessment Database 2020). Constrained by computational limits, MDP
approaches have been intractable on suitably realistic models and
largely confined to more academic applications (Costello et al. 2016).
However, advances in \emph{Deep Reinforcement Learning}, (DRL) a
sub-field of machine learning, have recently demonstrated remarkable
performance in a range of such MDP problems, from video games (Bellemare
et al. 2013; Mnih et al. 2013) to fusion reactions (Degrave et al. 2022;
Seo et al. 2022) to the remarkable dialog abilities of ChatGPT (OpenAI
2022). RL methods also bring many challenges of their own: being
notoriously difficult to train and evaluate, requiring immense
computational costs, and presenting frequent challenges with
reproducibility. A review of all these issues is beyond our scope but
can be found elsewhere (Lapeyrolerie et al. 2022; Chapman et al. 2023).
Here, though, we will focus on the issue of opacity and interpretability
raised by these methods. In contrast with optimization algorithms
currently used in either ecosystem management or resource economics, RL
algorithms have no guarantees of or metrics for convergence to an
optimal solution. In general, one can only assess the performance of
these black box methods relative to alternatives.

In most US fisheries, the mortality policy is often piecewise linear
(often with one constant and one linear piece), and the allowable
biological catch (ABC) or total allowable catch (TAC) is set at some
heuristic (e.g.~80\%) below the `overfishing limit', \(F_{MSY}\),
i.e.~the highest (constant) mortality that can be sustained indefinitely
(in the model -- reality of course does not permit such definitions).
This fixed mortality management can be seen, for instance, in most of
the fisheries listed in the widely used R.A. Myers Legacy Stock
Assessment Database. Here, we have focused on purely constant mortality
policies, rather than piecewise linear mortality funcitons, for
simplicity. Escapement-based management is less common, except in
salmonoids, as it requires closing a fishery whenever the measured
biomass falls below \(B_{MSY}\).

In this article, we compare against two common methods: constant
mortality (CMort) and constant escapement (CEsc), introduced in Text Box
1.\footnote{
A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘src’’ subdirectory. 
The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.
}

We consider the problem of devising harvest strategies in a for a series
of ecosystem models with increasing complexity (Table \ref{tab:models}):
\emph{1) one species, one fishery:} a simple single-species recruitment
model based on (May 1977); \emph{2) three species, one fishery:} a
three-species generalization of model \emph{1)}, where one of the
species is harvested; \emph{3) three species, two fisheries:} the same
three-species model as above but with two harvested species; \emph{4)
three species, two fisheries, parameter variation:} a three-species
model of which two are harvested, as above, with a time-varying
parameter. This last model is meant to be a toy model of climate
change's effect on the system. Across all of these scenarios, two goals
are balanced in the decision process: maximizing long-term catch and
preventing stock sizes to fall below some a-priori threshold.

\begin{table}
  \begin{center}
    \begin{tabular}{|l|c|c|c|c|}
      \hline
      \textbf{Model name} & \textbf{Model eqs.} & \textbf{N. Sp.} & \textbf{Harv. Sp.} & \textbf{Stationary?}\\
      \hline
      Model 1 & \eqref{eq:may} & 1 & $X$ & Yes \\
      Model 2 & \eqref{eq:3d model} & 3 & $X$ & Yes \\
      Model 3 & \eqref{eq:3d model} & 3 & $X$ and $Y$ & Yes \\
      Model 4 & \eqref{eq:3d model} & 3 & $X$ and $Y$ & No \\
      \hline
    \end{tabular}
    \caption{Table of models considered in this paper. Here, \textbf{N. Sp.} is the \emph{number of species} in the model, \textbf{Harv. Sp.} is the species of the model which are harvested, and \textbf{Stationary?} refers to whether the parameters of the model have fixed values (or, on the contrary, if they vary in time). The only non-stationary case presented in the paper is where $r_X$ drifts linearly with time. In the code repository associated to the paper, we consider other possible choices of non-stationarity.}
    \label{tab:models}
  \end{center}
\end{table}

This way, we evaluate classical management strategies (CMort and CEsc)
and DRL-based strategies on four different models. This experimental
design is summrized in Fig. \ref{fig:conceptual-v2}.

\begin{figure}[H]

{\centering \includegraphics[width=6.5in]{figures/conceptual-v3}

}

\caption{An experimental-design type of visualization of the management scenarios considered in this paper. On the x-axis are four different fishery management problems (Table 1). We represent the Model 4’s non-stationarity with a clock next to the X variable, and we intend to use it as an example of a possible simplified model for the effects of climate change. On the y-axis we have different management strategies with which one may control each of the models. On the bottom we have the constant escapement strategy (CEsc), based on calling off all fishing below a certain threshold population value. Above that is the constant mortality strategy (CMort), where one optimizes over constant fishing effort strategies. Finally, on top we have DRL-based strategies where policies are in general functions of the full state of the system, and they are parametrized by a neural network. The specific DRL-based strategy is referred to as PPO+GP in the main text, due to the algorithm used to produce the policy. The results plotted are the average reward obtained by the strategy over 100 episodes, and the fraction of those episodes which do not end with a near-extinction event (denoted Perc for Percentage). We have normalized to the highest reward in each column in order to enhance the comparison between strategies. For illustrative purposes we have color-coded the results using a two-dimensional color legend displayed on the bottom left.}\label{fig:conceptual-v2}
\end{figure}

Regarding control for these 4 models, we show the following. \emph{Model
1:} DRL-based strategies recover the optimal constant escapement (CEsc)
policy function. Constant mortality (CMort) performs considerably worse
than these three.\footnote{
As will be explained later, all our models are stochastic. 
If we set stochasticity to zero in Model 1, CMort matches the performance of the other management strategies.
} \emph{Model 2:} Here, all management strategies perform similarly.
\emph{Model 3:} For this model, DRL outperforms both classical
strategies, with CMort surprisingly performing significantly better than
CEsc. In particular, we observe that DRL strategies are more sensitive
to stochastic variations of the system which allows it to adaptively
manage the system to prevent population collapses below the threshold.
\emph{Model 4:} Here, the performance gap between DRL and both classical
strategies is maintained.

We show that in the most complex scenario, Model 4, CMort is faced with
a tradeoff---the optimal mortality rate leads a rather large fraction of
episodes ending with a population crash, whose negative reward is
counteracted with a higher economic output.\footnote{
In our mathematical formulation of the decision problem, we have assumed for simplicity that the fishing effort cost is zero and that fish price is stable over time.
This way, we equate economic output with harvested biomass.
} Conversely, more conservative, lower, mortality rates lead to lower
total reward on average. The DRL approach side-steps this trade-off by
optimizing over a more complex family of possible policies-----policies
parametrized by a neural network, as opposed to policies labeled by a
single parameter (the mortality rate).

Our findings paint a picture of how a single-species optimal management
strategy may lose performance rather dramatically when controlling a
more complex ecosystem. Here, DRL performs better from both economical
\emph{and} conservation points of view. Moreover, rather unintuitively,
CMort -- known to be suboptimal and unsustainable for single-species
models -- can turn out to even outperform the CEsc -- the single-species
optimal strategy -- for complex ecosystems. Finally, within this regime
of complex, possibly varying, ecosystems, we show that DRL consistently
finds a policy which effectively either matches the best classical
strategy, or outperforms it. We strengthen this result with a stability
investigation: we show that random perturbations in the model's
parameter values used do not significantly vary the conclusion that DRL
outperforms CEsc.

\hypertarget{mathematical-models-of-fisheries-considered}{%
\section{Mathematical models of fisheries
considered}\label{mathematical-models-of-fisheries-considered}}

Here we mathematically introduce the four fishery models for which we
compare different management strategies. In general, the class of models
that appear in this context are \emph{stochastic, first order, finite
difference equations}. For \(n\) species, these models have the general
form \begin{align}
  \label{eq:general model}
  \begin{split}
    \Delta X_t &= f_X(N_t) + \eta_{X,t} - M_X X_t\\
    \Delta Y_t &= f_Y(N_t) + \eta_{Y,t} - M_Y Y_t\\
    \Delta Z_t &= f_Z(N_t) + \eta_{Z,t} - M_Z Z_t\\
    &\dots
  \end{split}
\end{align} where
\(N_t = (X_t,\ Y_t,\ Z_t,\ \dots) \in \mathbb{R}^{n}_+\) is a vector of
populations, \(\Delta X_t := X_{t+1} - X_t\),
\(f_i:\mathbb{R}^{n}\to\mathbb{R}\) are arbitrary functions, and where
\(\eta_{i,t}\) are Gaussian random variables. Here,
\(M_i=M_i(N_t)\in\mathbb{R}_+\) is a state-dependent fish mortality
arising from harvesting the \(i\)-th species (sometimes this is referred
to as \emph{fishing effort}).

The term \(M_X X_t\) is the total \(X\) harvest at time \(t\). This
formulation of stock recruitment as a discrete finite difference process
is common among fisheries, as opposed to continuous time formulations
which involve instantaneous growth rates. This growth rate simplifies
e.g.~the possibly seasonal nature of reproduction (which would need to
be accounted for in a realistic continuous-time model) by simply
considering the total recruitment experienced by the population over a
full year.

The fishing efforts are the \emph{control variables} of our
problem---these may be set by the manager at each time-step to specified
values. We make two further simplifying assumptions on the control
problem: \emph{1. Full observation:} the manager is able to accurately
measure \(N_t\) and use that measurement to inform their decision.
\emph{2. Perfect execution:} the action chosen by the manager is
implemented perfectly (i.e., there is no noise affecting the actual
value of the fishing efforts).

Model 1 is a single-species classical model of ecological tipping
points. Models 2-4 are all three-species models with similar dynamics.
Following this logic, the first subsection will be dedicated to the
single-species model and the second will focus on the three-species
models.

\hypertarget{the-single-species-model}{%
\subsection{The single species model}\label{the-single-species-model}}

Optimal control policies for fisheries are frequently based on
1-dimensional models, \(n=1\), as described in \(Text\) \(Box\) \(1\).
The most familiar model of \(f(X)\) is that of \emph{logistic growth},
for which \begin{align}
  \label{eq:logistic}
  f(X_t) = r X_t\big(1 - X_t / K \big) =: L(X_t;\ r, K).
\end{align}

Real world ecological systems are obviously far more complicated than
this simple model suggests. One particularly important aspect that has
garnered much attention is the potential for the kind of highly
non-linear functions that can support dynamics such as alternative
stable states and hysteresis. A seminal example of such dynamics was
introduced in (May 1977), using a one-dimensional model of a prey
(resource) species under the pressure of a (fixed) predator. In the
notation of eq. \eqref{eq:general model}, \begin{align}
  \label{eq:may}
  f_X(X_t)
  = L(X_t;\ r, K) - \frac{\beta H X_t^2}{c^2 + X_t^2}.
\end{align} In the following, we will denote \begin{align*}
  F(X_t,H;\ \beta,c) := \frac{\beta H X_t^2}{c^2 + X_t^2}.
\end{align*}

The model has six parameters: the growth rate \(r\) and carrying
capacity \(K\) for \(X\), a constant population \(H\) of a species which
preys on \(X\), the maximal predation rate \(\beta\), the predation rate
half-maximum biomass \(c\), and the variance \(\sigma_X^2\) of the
stochastic term \(\eta_{X,t}\). (Here and in the following we will
center all random variables at zero.)

Eq. \eqref{eq:may} is an interesting study case of a \emph{tipping
point} (saddle-node bifurcation) (see Fig. \ref{fig:may}). Holding the
value of \(\beta\) fixed, for intermediate values of \(H\) there exist
two stable fixed points for the state \(X_t\) of the system, these two
attractors separated by an unstable fixed point. At a certain threshold
value of \(H\), however, the top stable fixed point collides with the
unstable fixed point and both are annihilated. For this value of \(H\),
and for higher values, only the lower fixed point remains. This also
creates the phenomenon of \emph{hysteresis}, where returning \(H\) to
its original value is not sufficient to restore \(X_t\) to the original
stable state.

\begin{figure}
\includegraphics[width=6.5in]{figures/MayStablePoints} \caption{The fixed point diagram for the unharvested dynamics of Model 1 as a function of varying the parameter $\beta H$, assuming zero noise. Stable fixed points (also known as attractors) are plotted using a solid line, while the unstable fixed point is shown as a dotted line.}\label{fig:may}
\end{figure}

This structure implies two things. First, that a drift in \(H\) could
lead to catastrophic consequences, with the population \(X_t\)
plummeting to the lower fixed stable point. Second, that if the
evolution of \(X_t\) is \emph{stochastic}, then, even at values of \(H\)
below the threshold point, the system runs a sizeable danger of tipping
over towards the lower stable point.

\hypertarget{the-three-species-models}{%
\subsection{The three species models}\label{the-three-species-models}}

Models 2-4 are three-species models and they are all closely
related---in fact, their natural dynamics (i.e.~dynamics under zero
harvest) is essentially given by the same equations: \begin{align}
\label{eq:3d model}
\begin{split}
  f_X(N_t) &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t, \\
  f_Y(N_t) &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
  f_Z(N_t) &= \left(b(X_t + DY_t\right) - d_Z)Z_t.
\end{split}
\end{align} The three species modeled are \(X\), \(Y\) and \(Z\).
Species \(Z\) preys on both \(X\) and \(Y\), while the latter two
compete for resources. There are thirteen parameters in this model: The
growth rate and carrying capacity, \(r_X\), \(K_X\), \(r_Y\) and
\(K_Y\), of \(X\) and \(Y\). A parameter \(c_{XY}\) mediating a
Lotka-Volterra competition between \(X\) and \(Y\). A maximum predation
rate \(\beta\) and a predation rate half-maximum biomass \(c\)
specifying how \(Z\) forages on \(X\) and \(Y\). A parameter \(D\)
regulating a relative preference of \(Z\) to prey on \(Y\). A death rate
\(d_Z\) and a parameter \(b\) scaling the birth rate of \(Z\). Finally,
the noise variances \(\sigma_X\), \(\sigma_Y\) and \(\sigma_Z\).

The three models will branch off of eq. \eqref{eq:3d model} in the
following way. \emph{Model 2:} here, only \(X\) is harvested, that is,
in the notation of eq. \eqref{eq:general model}, we fix \(M_Y=M_Z=0\)
and leave \(M_X\) as a control variable. All parameters here are
constant. \emph{Model 3:} as Model 1, but with \(X\) and \(Y\) being
harvested. In other words, we set \(M_Z=0\) and leave \(M_X\) and
\(M_Y\) as control variables. \emph{Model 4:} here \(X\) and \(Y\) are
harvested, but now we include a non-stationary parameter:
\begin{equation}
  \label{eq:rx(t)}
  r_X = r_X(t) = 
  \begin{cases}
    1-t/200, \quad & t \leq 100,\\
    1/2, \quad & t > 100.
  \end{cases}
\end{equation} All other parameters are constant. Eq. \eqref{eq:rx(t)}
is intended to reflect in a simple manner a possible effect of climate
change: where the reproductive rate of \(X\) is reduced linearly over
time until it stabilizes.

\hypertarget{sec:RL}{%
\section{Reinforcement learning}\label{sec:RL}}

Reinforcement learning (RL) is a way of approaching \emph{sequential
decision problems} through machine learning. All applications of RL can
be conceptually separated into two parts: an \emph{agent} and an
\emph{environment} which the agent interacts with. That is, the agent
performs actions within the environment.

After the agent takes an action, the environment will transition to a
new state and return a numerical \emph{reward} to the agent. (See Fig. 1
in (Lapeyrolerie et al. 2022) for a conceptual description of
reinforcement learning algorithms.) The rewards encode the agent's goal.
The main task of any RL algorithm is then to maximize the cumulative
reward received. This objective is achieved by aggregating experience in
what is called the \emph{training} period and learning from such
experience.

The \emph{environment} is commonly a computer simulation. It is
important to note here the role that real time-series data of stock
sizes can play in this process. This data is not used directly to train
the RL agent, but rather to estimate the model defining the environment.
This environment is subsequently used to train the agent. In this paper,
we focus on the second step---we take the estimated model of reality as
a given, and train an RL agent on it.\footnote{
In this sense, it is important to note that the classical management strategies we compare against have a similar flow of information.
Namely, data is used to estimate a dynamical model, and this model is used to generate a policy function.
The difference to our approach is located in the process of *how* the model is used to optimize a policy.
Because of this difference, RL-based approaches can produce good heuristic solutions for complex problems.
}

Specifically, we consider four environments corresponding to each of the
models considered (Table \ref{tab:models}). At each time step, the agent
observes the state \(S\) and enacts some harvest---reducing \(X_t\) to
\(X_t - M_X(N_t)\cdot X_t\), and, for Models 3 and 4, also reducing
\(Y_t\) to \(Y_t - M_Y(N_t)\cdot Y_t\). Here the fish
mortality-rates-from-harvest (i.e.~\(M_X=M_X(N_t)\) and
\(M_Y=M_Y(N_t)\)), are the agent's action at time \(t\). This secures a
reward of \(M_X(N_t)X\) for Models 1 and 2, and, similarly, a reward of
\(M_X(N_t)X + M_Y(N_t)Y\) for Models 3 and 4. After this harvest portion
of the time step, the environment evolves naturally according to eqs.
\eqref{eq:may} and \eqref{eq:3d model} (Sec. 2).

As mentioned previously, discretising time allows a simplification of
the possibly seasonal mating behavioral patterns of the species
involved. This approximation is commonly used in fisheries for species
with annual reproductive cycles (see e.g. (Mangel 2006), Chap. 6).
Moreover, the separation of each time-step into a harvest period and a
natural growth period assumes that harvest has little disruption on the
reproductive process. A detailed model which includes such a disruption
is outside of the scope of this work.

\hypertarget{mathematical-framework-for-rl}{%
\subsection{Mathematical framework for
RL}\label{mathematical-framework-for-rl}}

The RL environment can be formally described as a discrete time
\emph{partially observable Markov decision process (POMDP)}. This
formalization is rather flexible and allows one, e.g., to account for
situations where the agent may not fully observe the environment state,
or where the only observations available to the agent are certain
functions of the underlying state. For the sake of clarity, we will only
present here the subclass of POMDPs which are relevant to our work:
\emph{fully observable MDPs} (henceforth MDPs for short). An MDP may be
defined by the following data:

\begin{itemize}
\item
  \(\mathcal{S}\): \emph{state space}, the set of states of the
  environment,
\item
  \(\mathcal{A}\): \emph{action space}, the set of actions which the
  agent may choose from,
\item
  \(T(N_{t+1}|N_t, a_t, t)\): \emph{transition operator}, a conditional
  distribution which describes the dynamics of the system (where
  \(N_i\in\mathcal{S}\) are states of the environment),\footnote{
  Transition operators are commonly discussed without having a direct time-dependence for simplicity, but the inclusion of $t$ as an argument to $T$ does not alter the structure of the learning problem appreciably.
  }
\item
  \(r(N_t, a_t, t)\): \emph{reward function}, the reward obtained after
  performing action \(a_t\in\mathcal{A}\) in state \(N_t\),
\item
  \(d(N_0)\): \emph{initial state distribution}, the initial state of
  the environment is sampled from this distribution,
\item
  \(\gamma\in[0,1]\): \emph{discount factor}.
\end{itemize}

At a time \(t\), the MDP agent observes the full state \(s_t\) of the
environment and chooses an action based on this observation according to
a \emph{policy function} \(\pi(a_t | N_t)\). In return, it receives a
discounted reward \(\gamma^t r(a_t, N_t)\). The discount factor helps
regularize the agent, helping the optimization algorithm find solutions
which pay off within a timescale of \(t \sim \log(\gamma^{-1})^{-1}\).

With any fixed policy function, the agent will traverse a path
\(\tau=(N_0,\ a_0,\ N_1,\ a_1\ \dots,\ N_{t_{\text{fin.}}})\) sampled
randomly from the distribution \begin{align*}
  p_\pi(\tau) = 
  d(N_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
  \pi( a_t | N_t ) T( N_{t+1} | N_t, a_t, t ).
\end{align*} Reinforcement learning seeks to optimize \(\pi\) such that
the expected rewards are maximal, \[  
  \pi^* = \mathrm{argmax}\ \mathbb{E}_{\tau\sim p_\pi}[R(\tau)],
\] where, \[
  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, N_t, t),
\] is the cumulative reward of path \(\tau\). The function
\(J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]\) is called the
\emph{expected return}.

\hypertarget{deep-reinforcement-learning}{%
\subsection{Deep Reinforcement
Learning}\label{deep-reinforcement-learning}}
